---
title: "Supervised Learning"
author: "Andis Bara"
output: 
  html_document:
    theme: flatly
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
library(rsample)
library(recipes)
library(themis)
library(parsnip)
library(workflows)
library(tune)
library(doParallel)
library(ggplot2)
library(yardstick)
```

## Objective and Setup:
The goal of this supervised learning phase is to build predictive models that classify patients according to whether they are likely to experience a stroke based on their charachteristics and different factors.

The target outcome needs to be recognized as a classification factor
```{r}
df_clean <- as.data.frame(readRDS(here::here("data", "processed", "cleaned_stroke_dataset.rds")))

df_clean$stroke <- factor(df_clean$stroke,
                          levels = c("1", "0"),
                          labels = c("Stroke", "No Stroke"))

```

Isolating the Test Data and setting a seed

```{r}
set.seed(4092) 

# Stratified split that ensures the tiny percentage of stroke cases is distributed evenly
init_split <- initial_split(df_clean, prop = 0.80, strata = stroke)

train_df <- training(init_split)
test_df  <- testing(init_split)

levels(train_df$stroke)

# Printing to verify the imbalance in the training partition
cat("Original distribution in training data:\n")
print(table(train_df$stroke))
```
# Resampling the Training Data only:

### Why SMOTENC?
The dataset contains two highly unbalanced classes, stroke and no stroke. This can create problems when building predictive models, because the model may become biased toward the majority class and predict non-stroke cases most of the time, instead of learning patterns that identify the minority class properly. In this case, that would reduce the usefulness of the model.

To address this issue, a resampling method was needed to generate synthetic observations for the minority class. This gives the model more examples of stroke cases during training and helps it produce more meaningful predictions.

SMOTENC was chosen instead of SMOTE because the dataset includes both numerical and categorical variables. Since some predictors are categorical, SMOTENC is the more suitable method for the resampling step.

```{r}
smote_blueprint <- recipe(stroke ~ ., data = train_df) %>%
  # will be oversampled until it is 25% the size of the majority class.
  step_smotenc(stroke, over_ratio = 0.25) %>%
  step_dummy(all_nominal_predictors())

# Extracting the finalized, balanced dataset to use for model fitting
train_resampled <- smote_blueprint %>% 
  prep() %>% 
  juice()

cat("\nNew distribution after SMOTENC:\n")
print(table(train_resampled$stroke))
```

# Model Selection

For prediction I will be using: Logistic Regression, Random Forest & XGboosting.
-Why?

- Logistic Regression is used because it is one of the best methods for binary classification problems. It also allows for easier interpretability, which is important in this context.

- Random Forest was chosen because of it's ability to capture non-linear relationships better than Logistic Regression It is also more robust than a single decision tree and usually performs well in classification tasks.

- XGboost was included as a more advanced model, since it often gives strong predictive results on structured datasets.

### Preparing the models
- Logistic Regression spec
```{r}
log_spec <- logistic_reg() %>% 
  set_engine("glm") %>% 
  set_mode("classification")
```

- Random Forest spec:
We let it search for the optimal number of variables to check 
at each split (mtry) and the minimum node size (min_n).
```{r}
rf_tune_spec <- rand_forest(
  mtry = tune(),      
  min_n = tune(),     
  trees = 500         
) %>% 
  set_engine("ranger", importance = "impurity") %>% 
  set_mode("classification")
```

- XGboost spec:
We let it search for tree depth, learning rate, and node size.
```{r}
xgb_tune_spec <- boost_tree(
  tree_depth = tune(), 
  learn_rate = tune(), 
  min_n = tune(),
  trees = 500
) %>% 
  set_engine("xgboost") %>% 
  set_mode("classification")
```

### Cross-Validation Setup

Stratifying so the folds have a realistic number of stroke cases.
```{r}
set.seed(2026)

cv_folds <- vfold_cv(train_df, v = 5, strata = stroke)
cat("Successfully created 5 cross-validation folds.\n")
```

- Logistic Regression pipeline
```{r}
log_workflow <- workflow() %>%
  add_recipe(smote_blueprint) %>%
  add_model(log_spec)
```

- Random Forest pipeline
```{r}
rf_workflow <- workflow() %>%
  add_recipe(smote_blueprint) %>%
  add_model(rf_tune_spec)
```

- XGBoost pipeline
```{r}
xgb_workflow <- workflow() %>%
  add_recipe(smote_blueprint) %>%
  add_model(xgb_tune_spec)
```

# Parameter tuning

- Parallel processing is used to take advantage of multi-cores which allows for faster training time.

```{r parameter-tuning, cache = TRUE}
# Leaving one core free 
registerDoParallel(cores = parallel::detectCores() - 1) 

# Logistic regression doesn't need tuning, we just fit it directly to the folds
cat("Fitting Logistic Regression...\n")
log_results <- fit_resamples(
log_workflow,
resamples = cv_folds,
control = control_resamples(save_pred = TRUE)
)

# Tune Random Forest
cat("Tuning Random Forest...\n")
set.seed(345)
rf_tune_results <- tune_grid(
rf_workflow,
resamples = cv_folds,
grid = 10 
)

# Tune XGBoost
cat("Tuning XGBoost...\n")
set.seed(456)
xgb_tune_results <- tune_grid(
xgb_workflow,
resamples = cv_folds,
grid = 10 
)

# Shut down the parallel processing when done
stopImplicitCluster()
```

# Evaluating Model Performance

Now that the grid search is complete, we can visualize how the different hyperparameter combinations affected the performance of our tree-based models, and extract the optimal settings for each algorithm based on the Area Under the Curve (ROC-AUC).

```{r}
# Visualizing the tuning grids
autoplot(rf_tune_results) + 
  theme_minimal() + 
  labs(title = "Random Forest: Hyperparameter Tuning Results")

autoplot(xgb_tune_results) + 
  theme_minimal() + 
  labs(title = "XGBoost: Hyperparameter Tuning Results")

# Extracting the highest ROC-AUC scores for each model
cat(" Best Logistic Regression ---\n")
print(collect_metrics(log_results) %>% filter(.metric == "roc_auc"))

cat("\n Best Random Forest \n")
print(show_best(rf_tune_results, metric = "roc_auc", n = 1))

cat("\n Best XGBoost \n")
print(show_best(xgb_tune_results, metric = "roc_auc", n = 1))
```

```{r}
# Logistic Regression
set.seed(123)
log_compare_results <- fit_resamples(
  log_workflow,
  resamples = cv_folds,
  metrics = metric_set(accuracy, sens, spec)
)

# Random Forest
best_rf_params <- select_best(rf_tune_results, metric = "roc_auc")
final_rf_workflow <- finalize_workflow(rf_workflow, best_rf_params)

set.seed(123)
rf_compare_results <- fit_resamples(
  final_rf_workflow,
  resamples = cv_folds,
  metrics = metric_set(accuracy, sens, spec)
)

# XGBoost
best_xgb_params <- select_best(xgb_tune_results, metric = "roc_auc")
final_xgb_workflow <- finalize_workflow(xgb_workflow, best_xgb_params)

set.seed(123)
xgb_compare_results <- fit_resamples(
  final_xgb_workflow,
  resamples = cv_folds,
  metrics = metric_set(accuracy, sens, spec)
)

# Combine all three
compare_df <- bind_rows(
  collect_metrics(log_compare_results) %>% mutate(Model = "Logistic Regression"),
  collect_metrics(rf_compare_results) %>% mutate(Model = "Random Forest"),
  collect_metrics(xgb_compare_results) %>% mutate(Model = "XGBoost")
) %>%
  filter(.metric %in% c("accuracy", "sens", "spec")) %>%
  mutate(
    Metric = recode(.metric,
      accuracy = "Overall Accuracy",
      sens = "Sensitivity\n(Stroke)",
      spec = "Specificity\n(No Stroke)"
    )
  )

# Plot
ggplot(compare_df, aes(x = Metric, y = mean, fill = Model)) +
  geom_col(position = "dodge", color = "black", alpha = 0.85) +
  geom_text(
    aes(label = round(mean, 3)),
    position = position_dodge(width = 0.9),
    vjust = -0.5,
    size = 4
  ) +
  scale_fill_manual(values = c(
    "Logistic Regression" = "#e74c3c",
    "Random Forest" = "#2c3e50",
    "XGBoost" = "#27ae60"
  )) +
  theme_minimal() +
  labs(
    title = "Model Comparison",
    subtitle = "Cross-validated performance on the resampled training data",
    x = NULL,
    y = "Cross-Validated Score",
    fill = "Model"
  ) +
  coord_cartesian(ylim = c(0, 1.1))
```

### Interpretation:

- Even though the charts show that Random Forest and XGBoost were able to achieve better overall accuracy & and slightly better specificity, it is practically useless in terms of sensitivity with only 8.8% % 3% respectively "detection rate".

- On the other hand Logistic Regression was able to achieve a much higher sensitivity of 40.8% at the cost of a slight overall accuracy drop (3.9%) & slight specificity drop (5.8%).

- In the context of medical prediction, this is a major improvement because we are more willing to accept additional false positives in exchange for far fewer false negatives.

# Justifying the use of re-sampling for Logistic Regression

```{r}
# Creating a blueprint with no resampling simmilarly to the resampled one
unsampled_blueprint <- recipe(stroke ~ ., data = train_df) %>%
  step_dummy(all_nominal_predictors())

# Building a workflow using an unsampled Logistic Regression specification
unsampled_log_workflow <- workflow() %>%
  add_recipe(unsampled_blueprint) %>%
  add_model(log_spec)

# Fitting it to the exact same cross-validation folds to ensure a fair comparison
set.seed(999)
unsampled_log_results <- fit_resamples(
  unsampled_log_workflow,
  resamples = cv_folds,
  # collecting sensitivity and specifity
  metrics = metric_set(roc_auc, accuracy, sens, spec)
)

library(dplyr)
library(ggplot2)

# Combine sampled and unsampled Logistic Regression results
log_resampling_compare_df <- bind_rows(
  collect_metrics(log_compare_results) %>% mutate(Model = "Sampled (SMOTENC)"),
  collect_metrics(unsampled_log_results) %>% mutate(Model = "Unsampled (Raw Data)")
) %>%
  filter(.metric %in% c("accuracy", "sens", "spec")) %>%
  mutate(
    Metric = recode(.metric,
      accuracy = "Overall Accuracy",
      sens = "Sensitivity\n(Stroke)",
      spec = "Specificity\n(No Stroke)"
    )
  )

# Plot
ggplot(log_resampling_compare_df, aes(x = Metric, y = mean, fill = Model)) +
  geom_col(position = "dodge", color = "black", alpha = 0.85) +
  geom_text(
    aes(label = round(mean, 3)),
    position = position_dodge(width = 0.9),
    vjust = -0.5,
    size = 4
  ) +
  scale_fill_manual(values = c(
    "Sampled (SMOTENC)" = "#2c3e50",
    "Unsampled (Raw Data)" = "#e74c3c"
  )) +
  theme_minimal() +
  labs(
    title = "Logistic Regression: Unsampled vs. SMOTENC Resampling",
    subtitle = "Cross-validated performance before and after class balancing",
    x = NULL,
    y = "Cross-Validated Score",
    fill = "Training Data"
  ) +
  coord_cartesian(ylim = c(0, 1.1))
```

### Interpretation:

- This comparison shows why resampling was necessary. The unsampled model appeared to perform very well, with nearly perfect accuracy and specificity, but this result was misleading because it was mostly predicting “No Stroke,” the majority class, rather than learning to detect actual stroke cases.

```{r fig.width=10, fig.height=6}

# Fit logistic regression once on the training data
final_log_fit <- fit(log_workflow, data = train_df)

# Get predicted probabilities on the test set
log_probs <- predict(final_log_fit, test_df, type = "prob") %>%
  bind_cols(test_df %>% select(stroke))

# Thresholds to compare
thresholds_to_compare <- c(0.30, 0.20, 0.15)

# Build a table with accuracy, sensitivity, and specificity for each threshold
threshold_compare_df <- bind_rows(lapply(thresholds_to_compare, function(th) {
  
  preds <- log_probs %>%
    mutate(
      .pred_class = if_else(.pred_Stroke >= th, "Stroke", "No Stroke"),
      .pred_class = factor(.pred_class, levels = levels(stroke))
    )
  
  acc_df <- accuracy(preds, truth = stroke, estimate = .pred_class) %>%
    transmute(
      Threshold = paste0("t = ", th),
      Metric = "Overall Accuracy",
      Score = .estimate
    )
  
  sens_df <- sens(preds, truth = stroke, estimate = .pred_class) %>%
    transmute(
      Threshold = paste0("t = ", th),
      Metric = "Sensitivity\n(Stroke)",
      Score = .estimate
    )
  
  spec_df <- spec(preds, truth = stroke, estimate = .pred_class) %>%
    transmute(
      Threshold = paste0("t = ", th),
      Metric = "Specificity\n(No Stroke)",
      Score = .estimate
    )
  
  bind_rows(acc_df, sens_df, spec_df)
}))

# Plot side by side
ggplot(threshold_compare_df, aes(x = Metric, y = Score, fill = Threshold)) +
  geom_col(position = "dodge", color = "black", alpha = 0.85) +
  geom_text(
    aes(label = round(Score, 3)),
    position = position_dodge(width = 0.9),
    vjust = -0.5,
    size = 4
  ) +
  scale_fill_manual(values = c(
    "t = 0.3" = "#2c3e50",
    "t = 0.2" = "#e74c3c",
    "t = 0.15" = "#27ae60"
  )) +
  theme_minimal() +
  labs(
    title = "Logistic Regression Performance at Different Thresholds",
    subtitle = "Comparing overall accuracy, sensitivity, and specificity",
    x = NULL,
    y = "Performance Score",
    fill = "Threshold"
  ) +
  coord_cartesian(ylim = c(0, 1.1))
```

- This comparison shows that when the threshold is lowered, the model becomes better at catching stroke cases, since sensitivity goes up. At the same time, overall accuracy and specificity go down. So in the end, the threshold should be chosen depending on what is more important: keeping the model more balanced overall, or making sure that more stroke cases are detected.
